Gradual Soundness: Lessons from Static Python

### 2022-01-06 pj-7.1/paper.tex
pj: sec 5, use imports analysis to emphasize the mix of T-U code

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%% - is \sqle _consistent_ or _compatible_ subtyping?

@@

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
\cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
None of the converted modules have gotten slower because of \SP{}

\subsection{The Migration Path}

@@

During the first half of 2021, the \SP{} team identified critical
modules and proposed types to the maintainers of these modules.
Often, the types came with small code changes.
The accepted types and changes resulted in a 1\% improvement
in global CPU usage.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale by incorporating a tool that drafts
initial type annotations.
These changes resulted in a 2\% global CPU improvement.

As of December 2021, the \SP{} team has converted 530 modules.
Most of these came from the annotation tool (400); the rest
are from hand conversions.
In a typical week, the team adds types to four additional modules.
These modules have seen speedups ranging from 0\% to 25\% depending
on their nature.
Regressions have not been an issue.
The only cause for alarm is when a module gets only a small improvement
without code changes.
%% - rough measure: 0 to 25% per module
%%   ... but depends on nature of module
%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls

@@

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"

@@

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
\Cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
None of the converted modules have gotten slower because of \SP{}

\subsection{Migration Path}

@@

During the first half of 2021, the \SP{} team identified critical
modules and proposed types to the maintainers of these modules.
Often, the types came with small code changes.
The accepted types and changes resulted in a 1\% improvement
in global CPU usage.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale.
They also changed a code-generating module to output \SP{} code instead
of untyped Python code;
this one change brought over four hundred generated modules into the \SP{} fold.
Overall, these changes resulted in a 2\% global CPU improvement.

As of December 2021, the \SP{} team has converted 541 modules.
Most of these came from the tool (417); the rest
are from hand conversions (124).
These modules are well-mixed with untyped modules in the codebase.
All together, over thirty thousand identifiers cross a type boundary
at compile-time by either entering or leaving a \SP{} module.
Two-thirds of these crossings are exports from \SP{} modules
to untyped modules;
in other words, the typed identifiers are widely used throughout the Instagram web server.
%% data in src/static-python-imports-analysis

In a typical week, the team adds types to four additional modules.
These modules have seen speedups ranging from 0\% to 25\% depending
on their nature.
Regressions have not been an issue.
The only cause for alarm is when a module gets only a small improvement
without code changes.
%% - rough measure: 0 to 25% per module
%%   ... but depends on nature of module
%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls
%% 2022-01-06 FILL team did significant work to fight regressions

@@

%% 2022-01-06 TODO multiple dimensions of gradual
